---
title: "Visualizing ordinal regression models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing ordinal regression models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  eval = requireNamespace("ordinal", quietly = TRUE)
)
```

## Ordinal models as signal detection models

Rating-scale responses — "how true does this headline seem, from 1 to 6?" —
are ordinal: the categories are ordered, but the distances between them are
unknown. Cumulative link models (ordinal regression) handle this by assuming
a continuous latent variable that is cut into observed categories by a set
of thresholds. When the link is probit, this is exactly the model of classic
signal detection theory (SDT): each stimulus class produces a normal
distribution on a latent "perceived signal" axis, and the thresholds are the
response criteria. The distance between the distributions is the
discriminability (d'), and the placement of the thresholds captures response
bias.

This equivalence means an ordinal regression fit can be displayed in two
complementary ways. The latent-distribution plot shows the two stimulus
distributions and the thresholds directly. The receiver operating
characteristic (ROC) curve plots the hit rate against the false-alarm rate
at every threshold, summarizing discrimination independently of bias.
OrdinalRegressionViz draws both, for frequentist models fitted with
`ordinal::clm()`/`clmm()` and for Bayesian models fitted with `brms::brm()`.

These visualizations accompany Simchon, Zipori, Teitelbaum, Lewandowsky, and
van der Linden (2026), "A signal detection theory meta-analysis of
psychological inoculation against misinformation", *Current Opinion in
Psychology*, 67, 102194 (doi:10.1016/j.copsyc.2025.102194), where the same
displays are used to compare discrimination and response bias before and
after psychological inoculation interventions.

```{r setup}
library(OrdinalRegressionViz)
```

## The `sdt_ratings` dataset

The package ships with `sdt_ratings`, a simulated trial-level dataset in a
2 x 2 x 2 signal-detection design. Fifty participants in two conditions
(`control` / `inoculation`) rated true and fake news headlines (`target`)
before and after the intervention (`time`) on a 6-point perceived-truth
scale (`value`; 1 = "definitely fake", 6 = "definitely true"):

```{r data}
str(sdt_ratings)
```

The data were generated from an unequal-variance probit SDT model in which
inoculation increases post-intervention discrimination (d') and makes the
response criteria stricter. The generating code is in
`data-raw/sdt_ratings.R` in the source repository.

## Frequentist workflow

Fit a cumulative probit model on the control condition with
`ordinal::clm()`:

```{r clm-fit}
fit <- ordinal::clm(
  value ~ target * time,
  data = subset(sdt_ratings, condition == "control"),
  link = "probit"
)
```

### ROC curve

`roc_plot()` draws one ROC curve per level of `var_group`, with confidence
bands and one point per response threshold. The first level of `var_signal`
is treated as the signal class, and the hit rate is the probability of a low
response category given signal — here, giving a fake headline (`target =
"fake"`) a low perceived-truth rating:

```{r roc}
roc_plot(fit, var_signal = "target", var_group = "time")
```

The farther a curve bows away from the diagonal (chance performance), the
better participants discriminate fake from true headlines. In the control
condition the pre and post curves are close to each other, as expected: no
intervention took place between the two measurements.

Setting `show_empirical = TRUE` overlays the observed hit and false-alarm
rates as crosses — if the model fits, they should sit on or near the
model-implied curves:

```{r roc-empirical}
roc_plot(
  fit,
  var_signal = "target", var_group = "time",
  show_empirical = TRUE
)
```

### Latent distributions

`SDT_distributions_plot()` shows the same model as latent distributions.
Each panel corresponds to one level of `var_group` and displays the noise
and signal distributions together with the five estimated thresholds:

```{r distributions, fig.height = 7}
SDT_distributions_plot(fit, var_signal = "target", var_group = "time")
```

The horizontal distance between the two curves within a panel is d'. The
vertical lines are the response criteria: their spacing shows how the
6-point scale maps onto the latent axis, and shifts of the whole set between
panels indicate a change in response bias.

### Faceting by a third variable

To compare conditions, fit the model on the full dataset and pass the third
factor as `var_facet`. Two ROC panels are drawn side by side:

```{r roc-facet, fig.width = 8, fig.height = 5}
fit3 <- ordinal::clm(
  value ~ target * time * condition,
  data = sdt_ratings,
  link = "probit"
)

roc_plot(
  fit3,
  var_signal = "target", var_group = "time", var_facet = "condition"
)
```

In the inoculation panel the post curve bows out farther than the pre curve:
discrimination improved after the intervention. The same argument works in
`SDT_distributions_plot()`, which then draws a 2 x 2 grid of panels.

### SDT indices

`sdt_indices()` returns the quantities behind these displays as a tidy data
frame: the sensitivity index d', the signal/noise scale ratio, the response
criteria, and the area under the implied ROC curve, one set per design
cell:

```{r indices}
sdt_indices(
  fit3,
  var_signal = "target", var_group = "time", var_facet = "condition"
)
```

For `clm` models these are point estimates. The same call on a `brmsfit`
(below) uses the full posterior and adds credible intervals — useful when
you want to report, say, *d' = 1.9, 95% CrI [1.6, 2.2]*.

## Bayesian workflow

The Bayesian functions expect a cumulative model fitted with `brms::brm()`.
An unequal-variance SDT model adds a distributional part for `disc`, the
discrimination (inverse scale) of the latent distribution. Sampling takes a
few minutes, so the chunks in this section are not evaluated here:

```{r brm-fit, eval = FALSE}
b_fit <- brms::brm(
  brms::bf(value ~ target * time, disc ~ target * time),
  family = brms::cumulative("probit"),
  data = subset(sdt_ratings, condition == "control")
)
```

`bayesian_roc_plot()` draws the posterior ROC curve per level of
`var_group`, with curvewise credible bands:

```{r b-roc, eval = FALSE}
bayesian_roc_plot(b_fit, var_signal = "target", var_group = "time")
```

`bayesian_SDT_distribution_plot()` draws the posterior latent distributions
with credible bands around each density, and the posterior distribution of
each threshold as a slab on the x axis. With a `disc` part the two latent
distributions can have different widths (unequal-variance SDT); without it,
unit scales are used. `ndraws` subsamples the posterior draws to speed up
the density bands:

```{r b-dist, eval = FALSE}
bayesian_SDT_distribution_plot(
  b_fit,
  var_signal = "target", var_group = "time",
  plot_range = c(-4, 6), ndraws = 500
)
```

`ordinal_model_ppd_check()` compares the observed response distribution
with posterior predictive draws, in one facet per combination of the
grouping variables:

```{r b-ppd, eval = FALSE}
ordinal_model_ppd_check(b_fit, group_vars = c("target", "time"))
```

Finally, `sdt_indices()` on a `brmsfit` summarizes the posterior of every
SDT index with a credible interval:

```{r b-indices, eval = FALSE}
sdt_indices(b_fit, var_signal = "target", var_group = "time", ci = 0.95)
```

## Notes

- **Links other than probit.** The latent distribution follows the model's
  link function: normal for `probit`, logistic for `logit`, Cauchy for
  `cauchit`, and Gumbel for `cloglog`/`loglog`. The probit link gives the
  classic SDT display.
- **Saving plots.** All functions return `ggplot`/`patchwork` objects. Save
  them with `ggplot2::ggsave()`, for example
  `ggplot2::ggsave("roc.png", roc_plot(fit, "target", "time"), width = 7,
  height = 6, dpi = 300)`.
